---
title: "Sepsis and mortality prediction with GPT-5.2"
output: html_notebook
---

Here I will use GPT-5.2 + APACHE/SAPS to predict mortality. Limit to SepsisCx+ and SepsisCx- patients only.

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(pROC)

which_script <- "20260529b_ai_classifier.Rmd"
```

Define paths

```{r}
project_path <- if (basename(getwd()) == "code") ".." else "."
input_path <- file.path(project_path, "input")
```

ggplot theme

```{r}
my_theme <- theme(
  plot.title = element_text(hjust = 0.5, size=12, face="plain"),
  axis.text = element_text(size=12, color="black"),
  text = element_text(size=12, family="Arial"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.margin = unit(c(0.3,1,0.7,0), "cm")
)
```

Color palette

```{r}
# GPT only
color_gpt <- "#0173b2"

# Biomarker only
color_biomarker <- "#de8f05"

# Biomarker + GPT
color_biomarker_gpt <- "#029e73"
```

```{r}
sessionInfo()
```

## Import data

### Derivation cohort

Metadata file

```{r}
derivation_metadata <- read.csv(
  file.path(input_path, "metadata_derivationcohort.csv"),
  header=TRUE, row.names=NULL, check.names=TRUE
)

derivation_metadata <- derivation_metadata %>%
  select(Barcode, Sepsis, Hospital.Death, APACHEIII, SAPSII, gpt52_mortality_score) %>%
  mutate(
    across(c(Barcode), as.character),
    across(c(APACHEIII, SAPSII), as.numeric)) %>%
  filter(Sepsis %in% c("SepsisCx+", "SepsisCx-"))

print(dim(derivation_metadata))
print(head(derivation_metadata))
```

### Validation cohort

Metadata file

```{r}
validation_metadata <- read.csv(
  file.path(input_path, "metadata_validationcohort.csv"),
  header=TRUE, row.names=NULL, check.names=TRUE
)

validation_metadata <- validation_metadata %>%
  select(Barcode, Sepsis, Hospital.Death, APACHEIII, SAPSII, gpt52_mortality_score) %>%
  mutate(
    across(c(Barcode), as.character),
    across(c(APACHEIII, SAPSII), as.numeric)) %>%
  filter(Sepsis %in% c("SepsisCx+", "SepsisCx-"))

print(dim(validation_metadata))
print(head(validation_metadata))
```

## Sample preprocessing

Use the precomputed GPT-5.2 mortality scores in the public metadata files.

```{r}
derivation_gpt52_mortality_score <- derivation_metadata
print(head(derivation_gpt52_mortality_score))

validation_gpt52_mortality_score <- validation_metadata
print(head(validation_gpt52_mortality_score))
```

Order the cohort data by barcode.

```{r}
print(dim(derivation_gpt52_mortality_score))
derivation_gpt52_mortality_score <- derivation_gpt52_mortality_score %>%
  arrange(Barcode)
print(dim(derivation_gpt52_mortality_score))

print(dim(validation_gpt52_mortality_score))
validation_gpt52_mortality_score <- validation_gpt52_mortality_score %>%
  arrange(Barcode)
print(dim(validation_gpt52_mortality_score))
```

How many patients are there in each group?

```{r}
derivation_gpt52_mortality_score %>%
  dplyr::count(Hospital.Death) %>%
  print()

validation_gpt52_mortality_score %>%
  dplyr::count(Hospital.Death) %>%
  print()
```

In the derivation cohort, 6 patients have missing clinical data, and are excluded
from the metadata file. I'll exclude them here as well.

```{r}
print(dim(derivation_gpt52_mortality_score))

derivation_gpt52_mortality_score <- derivation_gpt52_mortality_score %>%
  filter(!is.na(Hospital.Death))

print(dim(derivation_gpt52_mortality_score))
```

## Mortality prediction: GPT-5.2 confusion matrix

```{r}
derivation_conf_matrix <- derivation_gpt52_mortality_score %>%
  mutate(
    gpt52_mortality_pred = if_else(gpt52_mortality_score>=1, "Mortality", "Survival")
  ) %>%
  dplyr::count(Hospital.Death, gpt52_mortality_pred)
print(derivation_conf_matrix)

validation_conf_matrix <- validation_gpt52_mortality_score %>%
  mutate(
    gpt52_mortality_pred = if_else(gpt52_mortality_score>=1, "Mortality", "Survival")
  ) %>%
  dplyr::count(Hospital.Death, gpt52_mortality_pred)
print(validation_conf_matrix)
```

```{r}
pa <- ggplot(
  data=derivation_conf_matrix,
  aes(x=gpt52_mortality_pred, y=Hospital.Death)) +
  geom_tile(aes(fill=n)) +
  geom_text(aes(label = n), size = 6, color="white") +
  scale_fill_viridis_c() +
  labs(
    title = "GPT-5.2 confusion matrix\nDerivation cohort",
    x = "Predicted",
    y = "Hospital mortality",
    fill = "Count"
  ) +
  theme_minimal() + my_theme + theme(aspect.ratio=1)

pb <- ggplot(
  data=validation_conf_matrix,
  aes(x=gpt52_mortality_pred, y=Hospital.Death)) +
  geom_tile(aes(fill=n)) +
  geom_text(aes(label = n), size = 6, color="white") +
  scale_fill_viridis_c() +
  labs(
    title = "GPT-5.2 confusion matrix\nValidation cohort",
    x = "Predicted",
    y = "Hospital mortality",
    fill = "Count"
  ) +
  theme_minimal() + my_theme + theme(aspect.ratio=1)

print(pa + pb)
```

## Mortality prediction: derivation cohort

I test how well the classifier performs in the derivation cohort with 5-fold CV.

### 5-fold CV

Generate stratified 5-fold CV

```{r}
derivation_gpt52_mortality_score %>%
  dplyr::count(Hospital.Death) %>%
  print()

set.seed(1)
folds <- derivation_gpt52_mortality_score %>%
  rsample::vfold_cv(
    v=5,
    strata=Hospital.Death)
```

ROC curve of GPT-5 alone

```{r}
pROC::roc(
  data = derivation_gpt52_mortality_score,
  response = Hospital.Death,
  predictor = gpt52_mortality_score,
  levels=c("No", "Yes"),
  direction="<",
  plot=TRUE, print.auc=TRUE
)
```

Calculate out-of-fold predicted probability of sepsis

```{r}
# For storing the out-of-fold probability
derivation_oof_prob <- list()

# Test 2 genes
my_biomarkers <- c("APACHEIII", "SAPSII")
counter <- 1

for (k in 1:length(folds$splits)) {
  train_folds <- rsample::analysis(folds$splits[[k]])
  test_fold <- rsample::assessment(folds$splits[[k]])
  
  # Convert Sepsis to factor for glm
  y_train <- factor(train_folds[["Hospital.Death"]], levels=c("No", "Yes"))
  y_test <- factor(test_fold[["Hospital.Death"]], levels=c("No", "Yes"))
  
  # --- GPT-5.2 only ---
  fit_0 <- glm(
    y_train ~ gpt52_mortality_score,
    data   = train_folds,
    family = "binomial"
  )
  prob_0 <- predict(
    fit_0,
    newdata = test_fold,
    type    = "response"
  )
  
  for (i in my_biomarkers) {
    # --- 1 biomarker ---
    fit_1 <- glm(
      y_train ~ x,
      data   = data.frame(x = train_folds[[i]]),
      family = "binomial",
      na.action = na.omit
    )
    
    # Predicted probability
    prob_1 <- predict(
      fit_1,
      newdata = data.frame(x = test_fold[[i]]),
      type    = "response",
      na.action = na.pass
    )
    
    # --- GPT-5.2 + 1 biomarker ---
    fit_2 <- glm(
      y_train ~ x + y,
      data   = data.frame(
        x = train_folds[[i]],
        y = train_folds$gpt52_mortality_score),
      family = "binomial",
      na.action = na.omit
    )
    
    # Predicted probability
    prob_2 <- predict(
      fit_2,
      newdata = data.frame(
        x = test_fold[[i]],
        y = test_fold$gpt52_mortality_score),
      type    = "response",
      na.action = na.pass
    )
    
    # --- Out-of-fold probabilities ---
    derivation_oof_prob[[counter]] <- data.frame(
      Barcode              = test_fold$Barcode,
      fold                 = k,
      biomarker            = i,
      y_true               = y_test,
      prob_gpt52           = prob_0,
      prob_biomarker       = prob_1,
      prob_biomarker_gpt52 = prob_2,
      stringsAsFactors     = FALSE
    )
    counter <- counter + 1
  }
}

# Prepare a data frame of out-of-fold probabilities
derivation_oof_prob <- bind_rows(derivation_oof_prob)
rownames(derivation_oof_prob) <- NULL  # clean up row names
```

### APACHEIII

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == "APACHEIII"),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == "APACHEIII"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == "APACHEIII"),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("APACHEIII AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("APACHEIII + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p1 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "APACHEIII only" = roc_biomarker,
    "APACHEIII + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Mortality classifier - Derivation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity",
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p1)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### SAPSII

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == "SAPSII"),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == "SAPSII"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == "SAPSII"),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("SAPSII AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("SAPSII + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p2 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "SAPSII only" = roc_biomarker,
    "SAPSII + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Mortality classifier - Derivation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity",
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p2)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### Save figure

```{r}
ggsave(
  "gpt52_clinical_mortality_roc_derivation.svg",
  plot = p1 + p2,
  height = 6, width = 10, units = "in"
)
```

## Mortality prediction: validation cohort

Train on derivation cohort, then test on validation cohort

```{r}
# For storing the out-of-fold probability
validation_prob <- list()

# Test 2 genes
my_biomarkers <- c("APACHEIII", "SAPSII")

# Convert mortality to factor for glm
y_train <- factor(
  derivation_gpt52_mortality_score[["Hospital.Death"]],
  levels=c("No", "Yes"))
y_test <- factor(
  validation_gpt52_mortality_score[["Hospital.Death"]],
  levels=c("No", "Yes"))

# --- GPT-5.2 only ---
fit_0 <- glm(
  y_train ~ gpt52_mortality_score,
  data   = derivation_gpt52_mortality_score,
  family = "binomial"
)
prob_0 <- predict(
  fit_0,
  newdata = validation_gpt52_mortality_score,
  type    = "response"
)

counter <- 1
for (i in my_biomarkers) {
  # --- 1 biomarker ---
  fit_1 <- glm(
    y_train ~ x,
    data   = data.frame(x = derivation_gpt52_mortality_score[[i]]),
    family = "binomial",
    na.action = na.omit
  )
  
  # Predicted probability
  prob_1 <- predict(
    fit_1,
    newdata = data.frame(x = validation_gpt52_mortality_score[[i]]),
    type    = "response",
    na.action = na.pass
  )
  
  # --- GPT-5.2 + 1 biomarker ---
  fit_2 <- glm(
    y_train ~ x + y,
    data   = data.frame(
      x = derivation_gpt52_mortality_score[[i]],
      y = derivation_gpt52_mortality_score$gpt52_mortality_score),
    family = "binomial",
    na.action = na.omit
  )
  
  # Predicted probability
  prob_2 <- predict(
    fit_2,
    newdata = data.frame(
      x = validation_gpt52_mortality_score[[i]],
      y = validation_gpt52_mortality_score$gpt52_mortality_score),
    type    = "response",
    na.action = na.pass
  )
  
  # --- Out-of-fold probabilities ---
  validation_prob[[counter]] <- data.frame(
    Barcode              = validation_gpt52_mortality_score$Barcode,
    biomarker            = i,
    y_true               = y_test,
    prob_gpt52           = prob_0,
    prob_biomarker       = prob_1,
    prob_biomarker_gpt52 = prob_2,
    stringsAsFactors     = FALSE
  )
  counter <- counter + 1
}


# Prepare a data frame of out-of-fold probabilities
validation_prob <- bind_rows(validation_prob)
rownames(validation_prob) <- NULL  # clean up row names
```

### APACHEIII

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == "APACHEIII"),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = validation_prob %>%
    filter(biomarker == "APACHEIII"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == "APACHEIII"),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("APACHEIII AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("APACHEIII + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p1 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "APACHEIII only" = roc_biomarker,
    "APACHEIII + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Mortality classifier - Validation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity",
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p1)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### SAPSII

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == "SAPSII"),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = validation_prob %>%
    filter(biomarker == "SAPSII"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == "SAPSII"),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No", "Yes"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("SAPSII AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("SAPSII + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p2 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "SAPSII only" = roc_biomarker,
    "SAPSII + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Mortality classifier - Validation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity",
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p2)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### Save figure

```{r}
ggsave(
  "gpt52_clinical_mortality_roc_validation.svg",
  plot = p1 + p2,
  height = 6, width = 10, units = "in"
)
```
