---
title: "Sepsis prediction with protein biomarkers"
output: html_notebook
---

Here I will use IL-6 and IL-8 to classify Sepsis vs No Sepsis.
- Derivation cohort: 5-fold cross-validation (out-of-fold predictions)
- Validation cohort: train on full derivation cohort, test on validation cohort

ZSFG patients are excluded from the derivation cohort (via the
Used_For_Prompt_Engineering filter). Only patients with available protein
data are kept.

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(pROC)

which_script <- "20260912_protein_sepsis_classifier.Rmd"
```

ggplot theme

```{r}
my_theme <- theme(
  plot.title    = element_text(hjust = 0.5, size = 12, face = "plain"),
  plot.subtitle = element_text(hjust = 0, size = 9, color = "grey30"),
  axis.text     = element_text(size = 12, color = "black"),
  text          = element_text(size = 12, family = "Arial"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.margin   = unit(c(0.3, 1, 0.7, 0), "cm")
)
```

Color palette

```{r}
color_il6 <- "#de8f05"   # orange  
color_il8 <- "#de8f05"   # teal
```

```{r}
sessionInfo()
```

## Import data

Protein data (contains both derivation and validation patients)

```{r}
protein <- read.csv(
  "input/20260429_derivation_validation_il6_il8.csv",
  header      = TRUE,
  row.names   = NULL,
  check.names = FALSE
)
protein <- protein %>%
  mutate(across(c(Barcode, EARLIStudyID), as.character))

print(dim(protein))
print(protein[1:6, ])
```

Derivation cohort metadata

```{r}
derivation_metadata <- read.csv(
  "../input/DerivationCohortMetadata052926.csv",
  header      = TRUE,
  row.names   = NULL,
  check.names = TRUE
)

derivation_metadata <- derivation_metadata %>%
  filter(Used_For_Prompt_Engineering == "No") %>%
  select(Barcode, EARLIStudyID, Sepsis, Hospital.Death, APACHEIII, SAPSII, Antibiotics) %>%
  mutate(across(c(Barcode, EARLIStudyID), as.character))

print(dim(derivation_metadata))
print(head(derivation_metadata))
```

Validation cohort metadata

```{r}
validation_metadata <- read.csv(
  "../input/ValidationCohortMetadata052926.csv",
  header      = TRUE,
  row.names   = NULL,
  check.names = TRUE
)

validation_metadata <- validation_metadata %>%
  select(Barcode, EARLIStudyID, Sepsis, Hospital.Death, APACHEIII, SAPSII, Antibiotics) %>%
  mutate(across(c(Barcode, EARLIStudyID), as.character))

print(dim(validation_metadata))
print(head(validation_metadata))
```

## Sample preprocessing

### Derivation cohort

Merge metadata with protein data; keep only patients with protein values

```{r}
derivation_data <- derivation_metadata %>%
  inner_join(protein %>% select(Barcode, il6, il8), by = "Barcode") %>%
  arrange(Barcode)

print(dim(derivation_data))
```

Filter to SepsisCx+ and No Sepsis only

```{r}
derivation_data %>% dplyr::count(Sepsis) %>% print()

derivation_data <- derivation_data %>%
  filter(Sepsis %in% c("SepsisCx+", "No Sepsis"))

derivation_data %>% dplyr::count(Sepsis) %>% print()
```

Drop patients with missing IL-6 or IL-8

```{r}
print(table(is.na(derivation_data$il6)))
print(table(is.na(derivation_data$il8)))

derivation_data <- derivation_data %>%
  filter(!is.na(il6), !is.na(il8))

print(dim(derivation_data))
```

Check for zeros before log-transform

```{r}
stopifnot(
  min(derivation_data$il6) > 0,
  min(derivation_data$il8) > 0
)
```

Log-transform

```{r}
derivation_data <- derivation_data %>%
  mutate(across(c(il6, il8), log10))
```

### Validation cohort

Merge metadata with protein data

```{r}
validation_data <- validation_metadata %>%
  inner_join(protein %>% select(Barcode, il6, il8), by = "Barcode") %>%
  arrange(Barcode)

print(dim(validation_data))
```

Filter to SepsisCx+ and No Sepsis only

```{r}
validation_data %>% dplyr::count(Sepsis) %>% print()

validation_data <- validation_data %>%
  filter(Sepsis %in% c("SepsisCx+", "No Sepsis"))

validation_data %>% dplyr::count(Sepsis) %>% print()
```

Drop patients with missing IL-6 or IL-8

```{r}
print(table(is.na(validation_data$il6)))
print(table(is.na(validation_data$il8)))

validation_data <- validation_data %>%
  filter(!is.na(il6), !is.na(il8))

print(dim(validation_data))
```

Check for zeros before log-transform

```{r}
stopifnot(
  min(validation_data$il6) > 0,
  min(validation_data$il8) > 0
)
```

Log-transform

```{r}
validation_data <- validation_data %>%
  mutate(across(c(il6, il8), log10))
```

## Sepsis classifier: derivation cohort (5-fold CV)

Generate stratified 5-fold CV

```{r}
set.seed(1)
folds <- derivation_data %>%
  rsample::vfold_cv(v = 5, strata = Sepsis)
```

Calculate out-of-fold predicted probabilities

```{r}
derivation_oof_prob <- list()
my_proteins <- c("il6", "il8")
counter <- 1

for (k in 1:length(folds$splits)) {
  train_folds <- rsample::analysis(folds$splits[[k]])
  test_fold   <- rsample::assessment(folds$splits[[k]])

  y_train <- factor(train_folds$Sepsis, levels = c("No Sepsis", "SepsisCx+"))
  y_test  <- factor(test_fold$Sepsis,   levels = c("No Sepsis", "SepsisCx+"))

  for (i in my_proteins) {
    fit <- glm(
      y_train ~ x,
      data   = data.frame(x = train_folds[[i]]),
      family = "binomial"
    )

    prob <- predict(
      fit,
      newdata = data.frame(x = test_fold[[i]]),
      type    = "response"
    )

    derivation_oof_prob[[counter]] <- data.frame(
      Barcode          = test_fold$Barcode,
      fold             = k,
      biomarker        = i,
      y_true           = y_test,
      prob_biomarker   = prob,
      stringsAsFactors = FALSE
    )
    counter <- counter + 1
  }
}

derivation_oof_prob <- bind_rows(derivation_oof_prob)
rownames(derivation_oof_prob) <- NULL
```

### IL-6 derivation cohort ROC

```{r}
roc_il6_deriv <- roc(
  data      = derivation_oof_prob %>% filter(biomarker == "il6"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

set.seed(1)
auc_il6_deriv <- ci.auc(roc_il6_deriv, method = "bootstrap", boot.n = 2000)
print(auc_il6_deriv)
```

### IL-8 derivation cohort ROC

```{r}
roc_il8_deriv <- roc(
  data      = derivation_oof_prob %>% filter(biomarker == "il8"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

set.seed(1)
auc_il8_deriv <- ci.auc(roc_il8_deriv, method = "bootstrap", boot.n = 2000)
print(auc_il8_deriv)
```

## Sepsis classifier: validation cohort (train on derivation, test on validation)

```{r}
validation_prob <- list()
counter <- 1

y_train <- factor(derivation_data$Sepsis, levels = c("No Sepsis", "SepsisCx+"))
y_test  <- factor(validation_data$Sepsis, levels = c("No Sepsis", "SepsisCx+"))

for (i in my_proteins) {
  fit <- glm(
    y_train ~ x,
    data   = data.frame(x = derivation_data[[i]]),
    family = "binomial"
  )

  prob <- predict(
    fit,
    newdata = data.frame(x = validation_data[[i]]),
    type    = "response"
  )

  validation_prob[[counter]] <- data.frame(
    Barcode          = validation_data$Barcode,
    biomarker        = i,
    y_true           = y_test,
    prob_biomarker   = prob,
    stringsAsFactors = FALSE
  )
  counter <- counter + 1
}

validation_prob <- bind_rows(validation_prob)
rownames(validation_prob) <- NULL
```

### IL-6 validation cohort ROC

```{r}
roc_il6_valid <- roc(
  data      = validation_prob %>% filter(biomarker == "il6"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

set.seed(1)
auc_il6_valid <- ci.auc(roc_il6_valid, method = "bootstrap", boot.n = 2000)
print(auc_il6_valid)
```

### IL-8 validation cohort ROC

```{r}
roc_il8_valid <- roc(
  data      = validation_prob %>% filter(biomarker == "il8"),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

set.seed(1)
auc_il8_valid <- ci.auc(roc_il8_valid, method = "bootstrap", boot.n = 2000)
print(auc_il8_valid)
```

## Plots and figure export

Helper function to build a single ROC panel

```{r}
make_roc_plot <- function(roc_obj, auc_ci, protein_label, cohort_label, color) {
  auc_annot <- sprintf(
    "%s AUC: %.2f (%.2f\u2013%.2f)",
    protein_label, auc_ci[2], auc_ci[1], auc_ci[3]
  )

  ggroc(
    roc_obj,
    legacy.axes = TRUE,
    linewidth   = 1,
    color       = color
  ) +
    geom_segment(
      aes(x = 0, xend = 1, y = 0, yend = 1),
      color    = "black",
      linetype = "dashed"
    ) +
    labs(
      title    = cohort_label,
      subtitle = auc_annot,
      x        = "1 \u2212 Specificity",
      y        = "Sensitivity"
    ) +
    theme_bw() + my_theme + theme(aspect.ratio = 1)
}
```

Build and preview the four panels

```{r}
pA <- make_roc_plot(roc_il6_deriv, auc_il6_deriv, "IL-6", "Derivation cohort", color_il6)
pB <- make_roc_plot(roc_il6_valid, auc_il6_valid, "IL-6", "Validation cohort", color_il6)
pC <- make_roc_plot(roc_il8_deriv, auc_il8_deriv, "IL-8", "Derivation cohort", color_il8)
pD <- make_roc_plot(roc_il8_valid, auc_il8_valid, "IL-8", "Validation cohort", color_il8)

print(pA)
print(pB)
print(pC)
print(pD)
```

Save individual panels for assembly in Illustrator

```{r}
ggsave("protein_sepsis_roc_il6_derivation.svg", plot = pA, height = 5, width = 5, units = "in")
ggsave("protein_sepsis_roc_il6_validation.svg",  plot = pB, height = 5, width = 5, units = "in")
ggsave("protein_sepsis_roc_il8_derivation.svg", plot = pC, height = 5, width = 5, units = "in")
ggsave("protein_sepsis_roc_il8_validation.svg",  plot = pD, height = 5, width = 5, units = "in")

cat("Saved 4 panels.\n")
```

## Summary table of AUC values

```{r}
auc_summary <- tribble(
  ~Protein, ~Cohort,      ~AUC,             ~CI_low,          ~CI_high,
  "IL-6",   "Derivation", auc_il6_deriv[2], auc_il6_deriv[1], auc_il6_deriv[3],
  "IL-6",   "Validation", auc_il6_valid[2], auc_il6_valid[1], auc_il6_valid[3],
  "IL-8",   "Derivation", auc_il8_deriv[2], auc_il8_deriv[1], auc_il8_deriv[3],
  "IL-8",   "Validation", auc_il8_valid[2], auc_il8_valid[1], auc_il8_valid[3]
) %>%
  mutate(across(c(AUC, CI_low, CI_high), ~round(., 3)))

print(auc_summary)
```
