---
title: "Sepsis and mortality prediction with GPT-5.2"
output: html_notebook
---

Here I will use GPT-5.2 + 1 gene biomarker to classify Sepsis vs No Sepsis. I remove ZSFG patients from the derivation cohort.

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)

library(pROC)
library(DESeq2)

which_script <- "20260529_ai_classifier.Rmd"
```

Define paths

```{r}
project_path <- if (basename(getwd()) == "code") ".." else "."
input_path <- file.path(project_path, "input")
```

ggplot theme

```{r}
my_theme <- theme(
  plot.title = element_text(hjust = 0.5, size=12, face="plain"),
  axis.text = element_text(size=12, color="black"),
  text = element_text(size=12, family="Arial"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.margin = unit(c(0.3,1,0.7,0), "cm")
)
```

Color palette

```{r}
# GPT only
color_gpt <- "#0173b2"

# Biomarker only
color_biomarker <- "#de8f05"

# Biomarker + GPT
color_biomarker_gpt <- "#029e73"
```

```{r}
sessionInfo()
```

## Utility functions

Take a train and test counts table, perform gene filter and VST gene normalization. Output: samples by genes.

```{r}
my_vst <- function(
    counts_train, counts_test,
    coldata_train, coldata_test) {
  
  # Train data gene normalization
  keep_train <- rowSums(edgeR::cpm(counts_train, log=FALSE)>=2) >= (0.2*ncol(counts_train))
  dds_train <- DESeqDataSetFromMatrix(
    countData = counts_train[keep_train,],
    colData   = coldata_train,
    design    = ~1)
  dds_train <- estimateSizeFactors(dds_train)
  dds_train <- estimateDispersions(dds_train)
  vsd_train <- vst(dds_train) %>%  # vst is faster than varianceStabilizingTransformation
    assay %>% 
    round(., digits=2) %>%
    t()
  
  # Test data gene normalization
  dds_test <- DESeqDataSetFromMatrix(
    countData = counts_test[keep_train,],
    colData   = coldata_test,
    design    = ~1)
  dds_test <- estimateSizeFactors(dds_test)
  dispersionFunction(dds_test) <- dispersionFunction(dds_train) # assign the dispersion function from the training data directly
  vsd_test <- vst(dds_test, blind=FALSE) %>%
    assay %>% 
    round(., digits=2) %>%
    t()
  
  return( list(vsd_train = vsd_train, vsd_test = vsd_test) )
}
```

## Import data

### Derivation cohort

Metadata file

```{r}
derivation_metadata <- read.csv(
  file.path(input_path, "metadata_derivationcohort.csv"),
  header=TRUE, row.names=NULL, check.names=TRUE
)

derivation_metadata <- derivation_metadata %>%
  select(Barcode, Sepsis, Sepsis_Original, gpt52_sepsis_score) %>%
  mutate(across(c(Barcode), as.character))

print(dim(derivation_metadata))
print(head(derivation_metadata))
```

Gene counts

```{r}
derivation_counts <- read.csv(
  file.path(input_path, "earli_counts_kallisto_derivation.csv"),
  row.names = 1
)
print(dim(derivation_counts))
print(derivation_counts[1:6,1:6])
```

### Validation cohort

Metadata file

```{r}
validation_metadata <- read.csv(
  file.path(input_path, "metadata_validationcohort.csv"),
  header=TRUE, row.names=NULL, check.names=TRUE
)

validation_metadata <- validation_metadata %>%
  select(Barcode, Sepsis, Sepsis_Original, gpt52_sepsis_score) %>%
  mutate(across(c(Barcode), as.character))

print(dim(validation_metadata))
print(head(validation_metadata))
```

Gene counts

```{r}
validation_counts <- read.csv(
  file.path(input_path, "earli_counts_kallisto_validation.csv"),
  row.names = 1, check.names = FALSE
)
print(dim(validation_counts))
print(validation_counts[1:6,1:6])
```

## Sample preprocessing

Verify that the two gene count matrices have the same genes

```{r}
stopifnot(all(rownames(derivation_counts) == rownames(validation_counts)))
```

Extract Barcode from the sample names of the gene count matrices

```{r}
colnames(derivation_counts) <- colnames(derivation_counts) %>%
  str_split_i("_", 2)
print(derivation_counts[1:6,1:6])

colnames(validation_counts) <- colnames(validation_counts) %>%
  str_split_i("_", 2)
print(validation_counts[1:6,1:6])
```

Patient 50577 has NA for every row. We will remove them.

```{r}
derivation_counts <- derivation_counts[,colnames(derivation_counts) != "50577"]
```

Use the precomputed GPT-5.2 sepsis scores in the public metadata files.

```{r}
derivation_gpt52_sepsis_score <- derivation_metadata
print(head(derivation_gpt52_sepsis_score))

validation_gpt52_sepsis_score <- validation_metadata
print(head(validation_gpt52_sepsis_score))
```

Only keep patients with available RNA-seq data.

```{r}
print(dim(derivation_gpt52_sepsis_score))
derivation_gpt52_sepsis_score <- derivation_gpt52_sepsis_score %>%
  filter(Barcode %in% colnames(derivation_counts)) %>%
  arrange(Barcode)
print(dim(derivation_gpt52_sepsis_score))

print(dim(validation_gpt52_sepsis_score))
validation_gpt52_sepsis_score <- validation_gpt52_sepsis_score %>%
  filter(Barcode %in% colnames(validation_counts)) %>%
  arrange(Barcode)
print(dim(validation_gpt52_sepsis_score))
```

How many patients are there in each group?

```{r}
derivation_gpt52_sepsis_score %>%
  dplyr::count(Sepsis) %>%
  print()

validation_gpt52_sepsis_score %>%
  dplyr::count(Sepsis) %>%
  print()
```

Filter both cohorts to SepsisCx+ and No Sepsis patients only

```{r}
derivation_gpt52_sepsis_score %>%
  dplyr::count(Sepsis) %>%
  print()
validation_gpt52_sepsis_score %>%
  dplyr::count(Sepsis) %>%
  print()

# Retain the broader cohorts for the supplemental subgroup analysis.
derivation_gpt52_sepsis_score_all <- derivation_gpt52_sepsis_score %>%
  filter(!is.na(Sepsis))
validation_gpt52_sepsis_score_all <- validation_gpt52_sepsis_score %>%
  filter(!is.na(Sepsis))

derivation_gpt52_sepsis_score <- derivation_gpt52_sepsis_score %>%
  filter(Sepsis %in% c("SepsisCx+", "No Sepsis"))
validation_gpt52_sepsis_score <- validation_gpt52_sepsis_score %>%
  filter(Sepsis %in% c("SepsisCx+", "No Sepsis"))

derivation_gpt52_sepsis_score %>%
  dplyr::count(Sepsis) %>%
  print()
validation_gpt52_sepsis_score %>%
  dplyr::count(Sepsis) %>%
  print()
```

## Sepsis diagnosis: GPT-5.2 confusion matrix

```{r}
derivation_conf_matrix <- derivation_gpt52_sepsis_score %>%
  filter(Sepsis %in% c("No Sepsis", "SepsisCx+")) %>%
  mutate(
    gpt52_sepsis_pred = if_else(gpt52_sepsis_score>=1, "Sepsis", "No Sepsis")
  ) %>%
  dplyr::count(Sepsis, gpt52_sepsis_pred)
print(derivation_conf_matrix)

validation_conf_matrix <- validation_gpt52_sepsis_score %>%
  filter(Sepsis %in% c("No Sepsis", "SepsisCx+")) %>%
  mutate(
    gpt52_sepsis_pred = if_else(gpt52_sepsis_score>=1, "Sepsis", "No Sepsis")
  ) %>%
  dplyr::count(Sepsis, gpt52_sepsis_pred)
print(validation_conf_matrix)
```

```{r}
pa <- ggplot(
  data=derivation_conf_matrix,
  aes(x=gpt52_sepsis_pred, y=Sepsis)) +
  geom_tile(aes(fill=n)) +
  geom_text(aes(label = n), size = 6, color="white") +
  scale_x_discrete(limits=c("Sepsis", "No Sepsis")) +
  scale_fill_viridis_c() +
  labs(
    title = "Derivation cohort",
    x = "Predicted",
    y = "True diagnosis",
    fill = "Count") +
  theme_minimal() + my_theme + theme(aspect.ratio=1)

pb <- ggplot(
  data=validation_conf_matrix,
  aes(x=gpt52_sepsis_pred, y=Sepsis)) +
  geom_tile(aes(fill=n)) +
  geom_text(aes(label = n), size = 6, color="white") +
  scale_x_discrete(limits=c("Sepsis", "No Sepsis")) +
  scale_fill_viridis_c() +
  labs(
    title = "Validation cohort",
    x = "Predicted",
    y = "True diagnosis",
    fill = "Count") +
  theme_minimal() + my_theme + theme(aspect.ratio=1)

print(pa + pb)
```

Save figure

```{r}
ggsave(
  "gpt52_sepsis_confusion_matrix.svg",
  plot = pa + pb,
  height = 4, width = 8, units = "in"
)
```

## Sepsis classifier: derivation cohort

I test how well the classifier performs in the derivation cohort with 5-fold CV.

### 5-fold CV

Generate stratified 5-fold CV

```{r}
set.seed(1)
folds <- derivation_gpt52_sepsis_score %>%
  rsample::vfold_cv(
    v=5,
    strata=Sepsis)
```

ROC curve of GPT-5 alone

```{r}
pROC::roc(
    Sepsis ~ gpt52_sepsis_score,
    data = derivation_gpt52_sepsis_score,
    levels=c("No Sepsis", "SepsisCx+"),
    direction="<",
    plot=TRUE, print.auc=TRUE
  )
```

Calculate out-of-fold predicted probability of sepsis

```{r}
# For storing the out-of-fold probability
derivation_oof_prob <- list()

# Test 2 genes
my_genes <- c(
  CD177 = "ENSG00000204936",
  MMP8  = "ENSG00000118113"
)
counter <- 1

for (k in 1:length(folds$splits)) {
  train_folds <- rsample::analysis(folds$splits[[k]])
  test_fold <- rsample::assessment(folds$splits[[k]])
  
  # Convert Sepsis to factor for glm
  y_train <- factor(train_folds$Sepsis, levels=c("No Sepsis", "SepsisCx+"))
  y_test <- factor(test_fold$Sepsis, levels=c("No Sepsis", "SepsisCx+"))
  
  # Gene normalization
  vsd <- my_vst(
    derivation_counts[,train_folds$Barcode],
    derivation_counts[,test_fold$Barcode],
    train_folds, test_fold
  )
  vsd_train <- vsd$vsd_train
  vsd_test <- vsd$vsd_test
  
  # --- GPT-5.2 only ---
  fit_0 <- glm(
    y_train ~ gpt52_sepsis_score,
    data   = train_folds,
    family = "binomial"
  )
  prob_0 <- predict(
    fit_0,
    newdata = test_fold,
    type    = "response"
  )
  
  for (i in my_genes) {
    # --- 1 biomarker ---
    fit_1 <- glm(
      y_train ~ x,
      data   = data.frame(x = vsd_train[,i]),
      family = "binomial"
    )
    
    # Predicted probability
    prob_1 <- predict(
      fit_1,
      newdata = data.frame(x = vsd_test[,i]),
      type    = "response"
    )
    
    # --- GPT-5.2 + 1 biomarker ---
    fit_2 <- glm(
      y_train ~ x + y,
      data   = data.frame(x = vsd_train[,i], y = train_folds$gpt52_sepsis_score),
      family = "binomial"
    )
    
    # Predicted probability
    prob_2 <- predict(
      fit_2,
      newdata = data.frame(x = vsd_test[,i], y = test_fold$gpt52_sepsis_score),
      type    = "response"
    )
    
    # --- Out-of-fold probabilities ---
    derivation_oof_prob[[counter]] <- data.frame(
      Barcode              = test_fold$Barcode,
      fold                 = k,
      biomarker            = i,
      y_true               = y_test,
      prob_gpt52           = prob_0,
      prob_biomarker       = prob_1,
      prob_biomarker_gpt52 = prob_2,
      stringsAsFactors     = FALSE
    )
    counter <- counter + 1
  }
}

# Prepare a data frame of out-of-fold probabilities
derivation_oof_prob <- bind_rows(derivation_oof_prob)
rownames(derivation_oof_prob) <- NULL  # clean up row names
```

### CD177

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("CD177 AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("CD177 + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p1 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "CD177 only" = roc_biomarker,
    "CD177 + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Sepsis classifier - Derivation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity", 
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p1)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### MMP8

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == my_genes["MMP8"]),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == my_genes["MMP8"]),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = derivation_oof_prob %>%
    filter(biomarker == my_genes["MMP8"]),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("MMP8 AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("MMP8 + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p2 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "MMP8 only" = roc_biomarker,
    "MMP8 + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Sepsis classifier - Derivation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity", 
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p2)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### Save figure

```{r}
ggsave(
  "gpt52_gene_sepsis_roc_derivation.svg",
  plot = p1 + p2,
  height = 6, width = 10, units = "in"
)
```

## Sepsis classifier: validation cohort

Train on derivation cohort, then test on validation cohort

```{r}
# For storing the predicted probability
validation_prob <- list()

# Test 2 genes
my_genes <- c(
  CD177 = "ENSG00000204936",
  MMP8  = "ENSG00000118113"
)

# Convert Sepsis to factor for glm
y_train <- factor(
  derivation_gpt52_sepsis_score$Sepsis,
  levels=c("No Sepsis", "SepsisCx+"))
y_test <- factor(
  validation_gpt52_sepsis_score$Sepsis,
  levels=c("No Sepsis", "SepsisCx+"))

# Gene normalization
vsd <- my_vst(
  derivation_counts[,derivation_gpt52_sepsis_score$Barcode],
  validation_counts[,validation_gpt52_sepsis_score$Barcode],
  derivation_gpt52_sepsis_score,
  validation_gpt52_sepsis_score
)
vsd_train <- vsd$vsd_train
vsd_test <- vsd$vsd_test

# --- GPT-5.2 only ---
fit_0 <- glm(
  y_train ~ gpt52_sepsis_score,
  data   = derivation_gpt52_sepsis_score,
  family = "binomial"
)
prob_0 <- predict(
  fit_0,
  newdata = validation_gpt52_sepsis_score,
  type    = "response"
)

counter <- 1
for (i in my_genes) {
  # --- 1 biomarker ---
  fit_1 <- glm(
    y_train ~ x,
    data   = data.frame(x = vsd_train[,i]),
    family = "binomial"
  )
  
  # Predicted probability
  prob_1 <- predict(
    fit_1,
    newdata = data.frame(x = vsd_test[,i]),
    type    = "response"
  )
  
  # --- GPT-5.2 + 1 biomarker ---
  fit_2 <- glm(
    y_train ~ x + y,
    data   = data.frame(
      x = vsd_train[,i],
      y = derivation_gpt52_sepsis_score$gpt52_sepsis_score),
    family = "binomial"
  )
  
  # Predicted probability
  prob_2 <- predict(
    fit_2,
    newdata = data.frame(
      x = vsd_test[,i],
      y = validation_gpt52_sepsis_score$gpt52_sepsis_score),
    type    = "response"
  )
  
  # --- Out-of-fold probabilities ---
  validation_prob[[counter]] <- data.frame(
    Barcode              = validation_gpt52_sepsis_score$Barcode,
    biomarker            = i,
    y_true               = y_test,
    prob_gpt52           = prob_0,
    prob_biomarker       = prob_1,
    prob_biomarker_gpt52 = prob_2,
    stringsAsFactors     = FALSE
  )
  counter <- counter + 1
}


# Prepare a data frame of out-of-fold probabilities
validation_prob <- bind_rows(validation_prob)
rownames(validation_prob) <- NULL  # clean up row names
```

### CD177

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("CD177 AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("CD177 + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p1 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "CD177 only" = roc_biomarker,
    "CD177 + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Sepsis classifier - Validation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity", 
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p1)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### MMP8

ROC curves

```{r}
roc_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["MMP8"]),
  response  = y_true,
  predictor = prob_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["MMP8"]),
  response  = y_true,
  predictor = prob_biomarker,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)

roc_biomarker_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["MMP8"]),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

Bootstrap AUC

```{r}
set.seed(1)
print(
  auc_gpt52 <- ci.auc(
    roc_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker <- ci.auc(
    roc_biomarker,
    method="bootstrap",
    boot.n=2000
  )
)

print(
  auc_biomarker_gpt52 <- ci.auc(
    roc_biomarker_gpt52,
    method="bootstrap",
    boot.n=2000
  )
)
```

Plot

```{r}
# Format AUC values
auc_annot <- glue::glue(
  sprintf("GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_gpt52[2], auc_gpt52[1], auc_gpt52[3]),
  "\n",
  sprintf("MMP8 AUC: %.3f (%.3f - %.3f)", auc_biomarker[2], auc_biomarker[1], auc_biomarker[3]),
  "\n",
  sprintf("MMP8 + GPT-5.2 AUC: %.3f (%.3f - %.3f)", auc_biomarker_gpt52[2], auc_biomarker_gpt52[1], auc_biomarker_gpt52[3]),
)

p2 <- ggroc(
  list(
    "GPT-5.2 only" = roc_gpt52,
    "MMP8 only" = roc_biomarker,
    "MMP8 + GPT-5.2" = roc_biomarker_gpt52),
  legacy.axes=TRUE,
  linewidth=1) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1), color="black", linetype="dashed") +
  scale_color_manual(values=c(color_gpt, color_biomarker, color_biomarker_gpt)) +
  labs(
    title = "Sepsis classifier - Validation cohort",
    subtitle = auc_annot,
    x = "1 \u2212 Specificity", 
    y = "Sensitivity",
    color = "Model") +
  theme_bw() + my_theme + theme(aspect.ratio=1)
print(p2)
```

Bootstrap test

```{r}
set.seed(1)
print(roc.test(
  roc_biomarker_gpt52, roc_gpt52,
  method = "bootstrap",
  boot.n = 2000
))

print(roc.test(
  roc_biomarker_gpt52, roc_biomarker,
  method = "bootstrap",
  boot.n = 2000
))
```

### Confusion matrix

What sensitivity, specificity and accuracy did the CD177 integrated classifier
achieve at Youden's index and at 95% sensitivity?

```{r}
roc_biomarker_gpt52 <- roc(
  data      = validation_prob %>%
    filter(biomarker == my_genes["CD177"]),
  response  = y_true,
  predictor = prob_biomarker_gpt52,
  levels    = c("No Sepsis", "SepsisCx+"),
  direction = "<",
  plot      = TRUE,
  print.auc = TRUE
)
```

What is the threshold at the Youden's index

```{r}
youden_thres <- coords(
  roc_biomarker_gpt52,
  "best",
  ret         = c("threshold", "sensitivity", "specificity", "accuracy"),
  best.method = "youden")
print(youden_thres)
```

Build the confusion matrix at Youden's index

```{r}
my_conf_matrix <- validation_prob %>%
  filter(
    biomarker == my_genes["CD177"]) %>%
  mutate(
    integrated_sepsis_pred = if_else(
      prob_biomarker_gpt52 >= youden_thres[["threshold"]],
      "Sepsis", "No Sepsis")
  ) %>%
  dplyr::count(y_true, integrated_sepsis_pred)
print(my_conf_matrix)
```

What is the threshold at 95% sensitivity

```{r}
sens95_thres <- coords(
  roc_biomarker_gpt52,
  "all",
  ret   = c("threshold", "sensitivity", "specificity", "accuracy")) %>%
  filter(sensitivity > 0.95) %>%
  filter(sensitivity == min(sensitivity)) %>%
  slice_max(specificity, n = 1, with_ties = FALSE)
print(sens95_thres)
```

Build the confusion matrix at this 95% sensitivity threshold.

```{r}
my_conf_matrix_2 <- validation_prob %>%
  filter(
    biomarker == my_genes["CD177"]) %>%
  mutate(
    integrated_sepsis_pred = if_else(
      prob_biomarker_gpt52 >= sens95_thres[["threshold"]],
      "Sepsis", "No Sepsis")
  ) %>%
  dplyr::count(y_true, integrated_sepsis_pred)
print(my_conf_matrix_2)
```

Plot the 2 confusion matrices


```{r}
pa <- ggplot(
  data=my_conf_matrix,
  aes(x=integrated_sepsis_pred, y=y_true)) +
  geom_tile(aes(fill=n)) +
  geom_text(aes(label = n), size = 6, color="white") +
  scale_x_discrete(limits=c("Sepsis", "No Sepsis")) +
  scale_fill_viridis_c() +
  labs(
    title = "Validation cohort: Youden's",
    x = "Predicted",
    y = "True diagnosis",
    fill = "Count") +
  theme_minimal() + my_theme + theme(aspect.ratio=1)

pb <- ggplot(
  data=my_conf_matrix_2,
  aes(x=integrated_sepsis_pred, y=y_true)) +
  geom_tile(aes(fill=n)) +
  geom_text(aes(label = n), size = 6, color="white") +
  scale_x_discrete(limits=c("Sepsis", "No Sepsis")) +
  scale_fill_viridis_c() +
  labs(
    title = "Validation cohort: 95% sens",
    x = "Predicted",
    y = "True diagnosis",
    fill = "Count") +
  theme_minimal() + my_theme + theme(aspect.ratio=1)

print(pa + pb)
```

### Save figures

```{r}
ggsave(
  "gpt52_gene_sepsis_roc_validation.svg",
  plot = p1 + p2,
  height = 6, width = 10, units = "in"
)
```

```{r}
ggsave(
  "integrated_sepsis_confusion_matrix_validation.svg",
  plot = pa + pb,
  height = 4, width = 8, units = "in"
)
```

## Supplemental analysis: classifier score by original Sepsis group

The derivation model is trained on SepsisCx+ and No Sepsis patients. It is then
applied to the remaining derivation patients and to the full validation cohort.

### Derivation holdout and full validation predictions

```{r}
holdout_data <- derivation_gpt52_sepsis_score_all %>%
  filter(!(Sepsis %in% c("SepsisCx+", "No Sepsis")))

supp_holdout_prob <- list()
y_train <- factor(
  derivation_gpt52_sepsis_score$Sepsis,
  levels=c("No Sepsis", "SepsisCx+"))
y_test <- holdout_data$Sepsis

vsd <- my_vst(
  derivation_counts[,derivation_gpt52_sepsis_score$Barcode],
  derivation_counts[,holdout_data$Barcode],
  derivation_gpt52_sepsis_score,
  holdout_data
)
vsd_train <- vsd$vsd_train
vsd_test <- vsd$vsd_test

fit_0 <- glm(
  y_train ~ gpt52_sepsis_score,
  data   = derivation_gpt52_sepsis_score,
  family = "binomial"
)
prob_0 <- predict(fit_0, newdata=holdout_data, type="response")

counter <- 1
for (i in my_genes) {
  fit_1 <- glm(
    y_train ~ x,
    data   = data.frame(x=vsd_train[,i]),
    family = "binomial"
  )
  prob_1 <- predict(
    fit_1,
    newdata=data.frame(x=vsd_test[,i]),
    type="response"
  )

  fit_2 <- glm(
    y_train ~ x + y,
    data=data.frame(
      x=vsd_train[,i],
      y=derivation_gpt52_sepsis_score$gpt52_sepsis_score),
    family="binomial"
  )
  prob_2 <- predict(
    fit_2,
    newdata=data.frame(
      x=vsd_test[,i],
      y=holdout_data$gpt52_sepsis_score),
    type="response"
  )

  supp_holdout_prob[[counter]] <- data.frame(
    Barcode              = holdout_data$Barcode,
    biomarker            = i,
    y_true               = y_test,
    prob_gpt52           = prob_0,
    prob_biomarker       = prob_1,
    prob_biomarker_gpt52 = prob_2,
    stringsAsFactors     = FALSE
  )
  counter <- counter + 1
}
supp_holdout_prob <- bind_rows(supp_holdout_prob)
rownames(supp_holdout_prob) <- NULL

supp_validation_prob <- list()
y_test <- validation_gpt52_sepsis_score_all$Sepsis

vsd <- my_vst(
  derivation_counts[,derivation_gpt52_sepsis_score$Barcode],
  validation_counts[,validation_gpt52_sepsis_score_all$Barcode],
  derivation_gpt52_sepsis_score,
  validation_gpt52_sepsis_score_all
)
vsd_train <- vsd$vsd_train
vsd_test <- vsd$vsd_test

prob_0 <- predict(
  fit_0,
  newdata=validation_gpt52_sepsis_score_all,
  type="response"
)

counter <- 1
for (i in my_genes) {
  fit_1 <- glm(
    y_train ~ x,
    data=data.frame(x=vsd_train[,i]),
    family="binomial"
  )
  prob_1 <- predict(
    fit_1,
    newdata=data.frame(x=vsd_test[,i]),
    type="response"
  )

  fit_2 <- glm(
    y_train ~ x + y,
    data=data.frame(
      x=vsd_train[,i],
      y=derivation_gpt52_sepsis_score$gpt52_sepsis_score),
    family="binomial"
  )
  prob_2 <- predict(
    fit_2,
    newdata=data.frame(
      x=vsd_test[,i],
      y=validation_gpt52_sepsis_score_all$gpt52_sepsis_score),
    type="response"
  )

  supp_validation_prob[[counter]] <- data.frame(
    Barcode              = validation_gpt52_sepsis_score_all$Barcode,
    biomarker            = i,
    y_true               = y_test,
    prob_gpt52           = prob_0,
    prob_biomarker       = prob_1,
    prob_biomarker_gpt52 = prob_2,
    stringsAsFactors     = FALSE
  )
  counter <- counter + 1
}
supp_validation_prob <- bind_rows(supp_validation_prob)
rownames(supp_validation_prob) <- NULL
```

### Supplemental figure

```{r}
supp_sepsis_dfs <- list(
  "derivation (OOF)" = derivation_oof_prob %>%
    left_join(
      derivation_metadata %>% select(Barcode, Sepsis_Original),
      by="Barcode"),
  "derivation (holdout)" = supp_holdout_prob %>%
    left_join(
      derivation_metadata %>% select(Barcode, Sepsis_Original),
      by="Barcode") %>%
    filter(Sepsis_Original != "Indeterminate"),
  "validation" = supp_validation_prob %>%
    left_join(
      validation_metadata %>% select(Barcode, Sepsis_Original),
      by="Barcode")
)

supp_sepsis_plots <- list()
counter <- 1
for (df in names(supp_sepsis_dfs)) {
  for (gene in names(my_genes)) {
    supp_sepsis_plots[[counter]] <- ggplot(
      data=supp_sepsis_dfs[[df]] %>%
        filter(biomarker == my_genes[gene]),
      aes(y=Sepsis_Original, x=prob_biomarker_gpt52)) +
      geom_boxplot(color="black", outlier.shape=NA, width=0.7) +
      ggbeeswarm::geom_quasirandom(size=0.7, orientation="y") +
      labs(
        y="Original Sepsis",
        x="Sepsis classifier score",
        title=paste0(df, " - ", gene)) +
      scale_x_continuous(limits=c(0,1.05), expand=c(0,0)) +
      theme_classic() + my_theme

    if (df == "derivation (OOF)") {
      supp_sepsis_plots[[counter]] <- supp_sepsis_plots[[counter]] +
        scale_y_discrete(
          limits=rev(c("SepsisBSI", "SepsisNonBSI", "No Sepsis")))
    } else if (df == "derivation (holdout)") {
      supp_sepsis_plots[[counter]] <- supp_sepsis_plots[[counter]] +
        scale_y_discrete(limits=rev(c("a", "SepsisCx-", "b")))
    } else if (df == "validation") {
      supp_sepsis_plots[[counter]] <- supp_sepsis_plots[[counter]] +
        scale_y_discrete(limits=rev(c(
          "SepsisBSI", "SepsisNonBSI", "SepsisCx-", "No Sepsis")))
    }
    counter <- counter + 1
  }
}

supp_sepsis_figure <- wrap_plots(supp_sepsis_plots, byrow=FALSE)
print(supp_sepsis_figure)

ggsave(
  "original_sepsis_sepsis.svg",
  plot=supp_sepsis_figure,
  height=5, width=10.7, units="in"
)
```

### Supplemental comparisons

```{r}
supp_sepsis_dfs[["validation"]] %>%
  dplyr::count(Sepsis_Original) %>%
  print()

for (i in names(my_genes)) {
  print(paste0("Gene: ", i))
  print(wilcox.test(
    prob_biomarker_gpt52 ~ as.factor(Sepsis_Original == "SepsisCx-"),
    data=supp_sepsis_dfs[["validation"]] %>%
      filter(Sepsis_Original != "No Sepsis", biomarker == my_genes[i])
  ))
}

for (i in names(my_genes)) {
  print(paste0("Gene: ", i))
  print(wilcox.test(
    prob_biomarker_gpt52 ~ as.factor(Sepsis_Original == "SepsisBSI"),
    data=supp_sepsis_dfs[["validation"]] %>%
      filter(
        Sepsis_Original %in% c("SepsisBSI", "SepsisNonBSI"),
        biomarker == my_genes[i])
  ))
}
```
