---
title: "Sepsis and mortality prediction with GPT-5.2"
output: html_notebook
---

Here I will use GPT-5.2 + 1 gene biomarker to classify Sepsis vs No Sepsis. I remove ZSFG patients from the derivation cohort.

```{r setup, include=FALSE}
library(tidyverse)

which_script <- "export_gpt52_score.Rmd"
```

Define paths

```{r}
project_path <- "/Users/natasha/Library/CloudStorage/Box-Box/20240903_LLM_sepsis_prediction"
output_path  <- "/Users/natasha/Library/CloudStorage/Box-Box/20240903_LLM_sepsis_prediction/github_repo/sepsis_GPT5_classifier/input"
```

## GPT-5.2 sepsis diagnosis

### Derivation cohort

GPT-5.2 sepsis diagnosis

```{r}
files <- Sys.glob(file.path(project_path, "Secure_out", "20260319_GPT5.2_sepsis_group*.csv"))
derivation_gpt52_sepsis <- bind_rows(lapply(files, read.csv))

derivation_gpt52_sepsis <- derivation_gpt52_sepsis %>%
  mutate(across(c(Barcode), as.character))
```

### Validation cohort

GPT-5.2 sepsis diagnosis

```{r}
files <- Sys.glob(file.path(project_path, "Secure_out", "20260330_GPT5.2_sepsis_group*.csv"))
validation_gpt52_sepsis <- bind_rows(lapply(files, read.csv)) %>%
  mutate(Barcode = as.character(Barcode))
```

### Aggregate sepsis score

Verify that each patient has 10 GPT-5.2 responses

```{r}
derivation_gpt52_sepsis %>%
  dplyr::count(Barcode) %>%
  filter(n != 10) %>%
  pull(Barcode) %>%
  print()

validation_gpt52_sepsis %>%
  dplyr::count(Barcode) %>%
  filter(n != 10) %>%
  pull(Barcode) %>%
  print()
```

Aggregate the GPT-5.2 sepsis responses into the sepsis score

```{r}
derivation_gpt52_sepsis_score <- derivation_gpt52_sepsis %>%
  group_by(Barcode) %>%
  summarize(gpt52_sepsis_score = sum(sepsis=="Sepsis"))
print(head(derivation_gpt52_sepsis_score))

validation_gpt52_sepsis_score <- validation_gpt52_sepsis %>%
  group_by(Barcode) %>%
  summarize(gpt52_sepsis_score = sum(sepsis=="Sepsis")) %>%
  mutate(Barcode = str_replace_all(Barcode, "_", "-"))
print(head(validation_gpt52_sepsis_score))
```

## GPT-5.2 mortality prediction

### Derivation cohort

```{r}
files <- Sys.glob(file.path(project_path, "Secure_out", "20260407_GPT5.2_mortality_group*.csv"))
derivation_gpt52_mortality <- bind_rows(lapply(files, read.csv))

derivation_gpt52_mortality <- derivation_gpt52_mortality %>%
  mutate(across(c(Barcode), as.character))
```

### Validation cohort

```{r}
files <- Sys.glob(file.path(project_path, "Secure_out", "20260408_GPT5.2_mortality_group*.csv"))
validation_gpt52_mortality <- bind_rows(lapply(files, read.csv)) %>%
  mutate(Barcode = as.character(Barcode))
```

### Aggregate mortality score

Verify that each patient has 10 GPT-5.2 responses

```{r}
derivation_gpt52_mortality %>%
  dplyr::count(Barcode) %>%
  filter(n != 10) %>%
  pull(Barcode) %>%
  print()

validation_gpt52_mortality %>%
  dplyr::count(Barcode) %>%
  filter(n != 10) %>%
  pull(Barcode) %>%
  print()
```

Aggregate

```{r}
derivation_gpt52_mortality_score <- derivation_gpt52_mortality %>%
  group_by(Barcode) %>%
  summarize(gpt52_mortality_score = sum(mortality=="Not survive"))
print(head(derivation_gpt52_mortality_score))

validation_gpt52_mortality_score <- validation_gpt52_mortality %>%
  group_by(Barcode) %>%
  summarize(gpt52_mortality_score = sum(mortality=="Not survive")) %>%
  mutate(Barcode = str_replace_all(Barcode, "_", "-"))
print(head(validation_gpt52_mortality_score))
```

## Load metadata

```{r}
derivation_metadata <- read.csv(
  "/Users/natasha/Library/CloudStorage/Box-Box/20240903_LLM_sepsis_prediction/github_repo/sepsis_GPT5_classifier/input/DerivationCohortMetadata052926.csv",
  header = TRUE, row.names = NULL, check.names = TRUE
)

validation_metadata <- read.csv(
  "/Users/natasha/Library/CloudStorage/Box-Box/20240903_LLM_sepsis_prediction/github_repo/sepsis_GPT5_classifier/input/ValidationCohortMetadata052926.csv",
  header = TRUE, row.names = NULL, check.names = TRUE
)
```

## Sanity check column names

```{r}
cat("Derivation metadata columns:\n")
print(names(derivation_metadata))

cat("\nValidation metadata columns:\n")
print(names(validation_metadata))
```

Confirm expected columns exist before proceeding

```{r}
deriv_expected <- c("Barcode", "Age", "Gender", "Sepsis", "Sepsis_Original",
                    "Hospital.Death", "APACHEIII", "SAPSII", "Antibiotics",
                    "Used_For_Prompt_Engineering")

valid_expected <- c("Barcode", "Age", "Gender", "Sepsis", "Sepsis_Original",
                    "Hospital.Death", "APACHEIII", "SAPSII", "Antibiotics")

missing_deriv <- setdiff(deriv_expected, names(derivation_metadata))
missing_valid <- setdiff(valid_expected, names(validation_metadata))

if (length(missing_deriv) > 0) {
  cat("WARNING: Missing derivation columns:", paste(missing_deriv, collapse=", "), "\n")
} else {
  cat("Derivation columns OK\n")
}

if (length(missing_valid) > 0) {
  cat("WARNING: Missing validation columns:", paste(missing_valid, collapse=", "), "\n")
} else {
  cat("Validation columns OK\n")
}
```

## Build final output files

Cap age at 89 for HIPAA compliance (ages >89 are set to 89)

```{r}
derivation_metadata <- derivation_metadata %>%
  mutate(Age = pmin(as.numeric(Age), 89))

validation_metadata <- validation_metadata %>%
  mutate(Age = pmin(as.numeric(Age), 89))

cat("Derivation ages >89 after capping:", sum(derivation_metadata$Age > 89, na.rm=TRUE), "\n")
cat("Validation ages >89 after capping:", sum(validation_metadata$Age > 89, na.rm=TRUE), "\n")
```

Derivation cohort: metadata + sepsis score + mortality score

```{r}
derivation_out <- derivation_metadata %>%
  mutate(Barcode = as.character(Barcode)) %>%
  select(all_of(deriv_expected)) %>%
  left_join(derivation_gpt52_sepsis_score,    by = "Barcode") %>%
  left_join(derivation_gpt52_mortality_score,  by = "Barcode")

cat("Derivation output dimensions:", nrow(derivation_out), "x", ncol(derivation_out), "\n")
print(head(derivation_out))
```

Validation cohort: metadata + sepsis score + mortality score

```{r}
validation_out <- validation_metadata %>%
  mutate(Barcode = as.character(Barcode)) %>%
  select(all_of(valid_expected)) %>%
  left_join(validation_gpt52_sepsis_score,    by = "Barcode") %>%
  left_join(validation_gpt52_mortality_score,  by = "Barcode")

cat("Validation output dimensions:", nrow(validation_out), "x", ncol(validation_out), "\n")
print(head(validation_out))
```

## Sanity checks

Check for patients with missing GPT scores (did not match on Barcode)

```{r}
cat("Derivation patients missing sepsis score:",
    sum(is.na(derivation_out$gpt52_sepsis_score)), "\n")
cat("Derivation patients missing mortality score:",
    sum(is.na(derivation_out$gpt52_mortality_score)), "\n")
cat("Validation patients missing sepsis score:",
    sum(is.na(validation_out$gpt52_sepsis_score)), "\n")
cat("Validation patients missing mortality score:",
    sum(is.na(validation_out$gpt52_mortality_score)), "\n")
```

Score distributions (should be integers 0-10)

```{r}
cat("Derivation sepsis score range:",
    min(derivation_out$gpt52_sepsis_score, na.rm=TRUE), "-",
    max(derivation_out$gpt52_sepsis_score, na.rm=TRUE), "\n")
cat("Derivation mortality score range:",
    min(derivation_out$gpt52_mortality_score, na.rm=TRUE), "-",
    max(derivation_out$gpt52_mortality_score, na.rm=TRUE), "\n")
cat("Validation sepsis score range:",
    min(validation_out$gpt52_sepsis_score, na.rm=TRUE), "-",
    max(validation_out$gpt52_sepsis_score, na.rm=TRUE), "\n")
cat("Validation mortality score range:",
    min(validation_out$gpt52_mortality_score, na.rm=TRUE), "-",
    max(validation_out$gpt52_mortality_score, na.rm=TRUE), "\n")
```

## Export to input/ folder

```{r}
write.csv(
  derivation_out,
  file.path(output_path, "metadata_derivationcohort.csv"),
  row.names = FALSE
)

write.csv(
  validation_out,
  file.path(output_path, "metadata_validationcohort.csv"),
  row.names = FALSE
)

cat("Files written to:", output_path, "\n")
cat("Rows written - Derivation:", nrow(derivation_out),
    "| Validation:", nrow(validation_out), "\n")
```
